---
title: "Creating Landscape Fragmentation Statistics"
author: "Michael Fichman & Ken Steif"
date: "12/29/2021"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: "hide"
    code_download: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Intro

blah blah

Excellent tutorial for dealing with raster data in R

https://www.emilyburchfield.org/courses/eds/rasters_in_r

## Libraries

```{r cars}
library(tidyverse)
library(raster)
library(sf)
library(mapview)
```

## Load Data

 Load 2001 and 2011 Chesco Landcover Data

```{r load_data}
cc01 <- raster("~/GitHub/CPLN_675/Week_2/data/R_Data/cc01.tif")
cc11 <- raster("~/GitHub/CPLN_675/Week_2/data/R_Data/cc11.tif")
```
## Examine the data

What is the cell size? 

What is the projection? 

What is the extent?

```{r}
plot(cc01)

plot(cc11)
```

## Reclassify
Reclassify using a matrix - e.g. if between col1 and col2, value = col3

11 - Open Water
12 - Perennial Snow/Ice
21 - Developed - Open Space
22 - Low Density - Developed
23 - Moderate Density - Developed
24 - High Density - Developed
41 - Deciduous Forest
42 - Evergreen Forest
43 - Mixed Forest
52 - Shrub/Scrub
71 - Herbacious
81 - Hay/pasture
82 - Cultivated crops
90 - Woody  Wetland
95 - Emergent Herbacious Wetland

```{r reclass}
reclass_df <-c(0, 10, NA,
               10, 12, 1,
               20, 24, 0,
               25, Inf, 1)
               
reclass_matrix <-  matrix(reclass_df, ncol = 3,
                byrow = TRUE)

cc01_rc <- reclassify(cc01,
                     reclass_matrix)

cc11_rc <- reclassify(cc11,
                     reclass_matrix)
```

## Load roads

```{r}
roads <- st_read("~/GitHub/CPLN_675/Week_2/data/R_Data/chesco_roads.geojson")
```

Go feature to raster

rasterize(vector (turned into sp spatial object), reference raster)

REMOVE THE SLICE WHEN THIS GOES LIVE

```{r}
roads_raster <- rasterize(roads %>%
                           # slice(1:100) %>%
                            mutate(road = 1) %>%
                            dplyr::select(road) %>%
                            as(., "Spatial"),
                          cc11)

reclass_df <-c(1, nrow(roads), 0,
               NA, NA, 1)
               
reclass_matrix <-  matrix(reclass_df, ncol = 3,
                byrow = TRUE)

roads_rc <- reclassify(roads_raster,
                     reclass_matrix)
```


## Combine Roads with Reclassified Raster

cc11_rc - developed == 0
roads_rc - developed = 0

```{r}
cc11_rc_roads <- cc11_rc * roads_rc

plot(cc11_rc_roads)
```

## Crop the data by the Chesco shapefile

Go feature to raster on the Chesco shapefile and then raster calculator to get rid of stuff outside the boundary

## Region Group




OLD ROUTINE

Read in csvs for 2001 and 2011 land use data. You will have to find the appropriate filepath for where these data are stored on your machine.

```{r read_data, message=FALSE, warning=FALSE, include=TRUE}
a <- read.csv("~/Clients/MUSA_Teaching_and_Admin/CPLN_675_2020/wk_2/data/z_extra/2001.txt")
b <- read.csv("~/Clients/MUSA_Teaching_and_Admin/CPLN_675_2020/wk_2/data/z_extra/2011.txt")
```

# 3. Explore and clean data

Let's look at our new data frames using some commands which ask about the nature or dimensions of our data.

Try some of these functions in your console window:

```{r explore_data, eval=FALSE, message=FALSE, warning=FALSE}
class(a)
names(a)
nrow(a)
ncol(a)
str(a)
```

My personal favorite is `glimpse()` - it shows you most of what you need to know.

```{r glimpse_data, echo=TRUE}
glimpse(a)
```

We will create two new dataframes which contain only the variables we need. We'll pull out only what we want - the GRIDCODE (total not developed patches), AREA, perimeter, iso and MUNI_NAM (town)

There are several ways we can do this. Often in programming, you use column index numbers and pull out variables using matrix notation. That would look like this:

`a2 <- a[,c(5,6,7,8,12)]`

With the `tidyverse` and `dplyr`, you can use the much simpler `select` command and subset the columns into a new dataframe - called `a2` - by name.

Notice we "pipe" the `select` command to dataframe `a` using the `%>%` operator (aka the pipe) which allows you to chain functions together. In plain English, here we are saying "From dataframe a, select the following columns and call that data frame a2"

```{r create_a2, echo=TRUE}
a2 <- a %>%
  select(gridcode, area, perimeter, iso, MUNI_NAME)
```

Let's use the `head` function to look at the first 5 rows of our new data frame

```{r head_a, echo=FALSE}
head(a2)
```

We want to be able to append 2001 and 2011 data together and summarize them by year. In order to do this, both data frames need to have the same variable names. We are going to have to create a year variable to differentiate our data once both data frames are appended. We do so for each using the same column name.

I also want to create a column called `year` which indicates that these are my 2001 data. I'm going to put them together later and I'll need to know which data are which. I can add that data using the `mutate` command from `dplyr`. In plain English, the operation here is as follows - "To data frame a2, mutate a new column called year and set it equal to the text (aka character) string "2001". Overwrite a2 with this new data frame".

```{r create_year, echo=TRUE}
a2 <- a2 %>%
  mutate(year = "2001")
```

We can do the same for data frame `b`. Notice here, I will add the year column like I did to `a2` but within the sequence of piped commands.

```{r create_var_b, echo=TRUE}
b2 <- b %>%
  select(gridcode, area, perimeter, iso, MUNI_NAME) %>%
  mutate(year = "2011")
```

We can use a boolean expression to see if our names are the same, or we can use the function `setdiff` to figure out what might be different

```{r variable_check, echo=TRUE}
names(a2) == names(b2)
setdiff(names(a2), names(b2))
```

Now we can append these data frames together by row using `rbind` ("row bind"). You can put two data frames together in this manner if they have the same number of columns and column names and types.

Just to make sure this `rbind` went off without a hitch we will create a table to see if both years are present in the new data frame.

```{r bind_years, echo=TRUE}
data <- rbind(a2,b2)

table(data$year)
```

# 4. Summarize Data

Let's use some tidy functions to `summarize` our data by municipality and by year and then `filter` out any observations that are not in a municipality.

Here we `group_by` year and municipality and summarize some patch statistics.

```{r summarize, echo=TRUE, message = FALSE, warning = FALSE}
summary <- data %>%
  group_by(year, MUNI_NAME) %>%
  summarize(countPatch = sum(gridcode),                      #count of unique patches
            mean_Compactness = mean(iso),                    #mean 'roundness' of patch 
            mean_Area = mean(area),                          #mean area of patch 
            sum_Area = sum(area),                            #sum area of patch
            sum_Perimter = sum(perimeter)) %>%
  filter(MUNI_NAME != " ") %>%
  as.data.frame()
```

Now that we have summarized by municipality and year, we can calculate change for each town. Ultimately we are going to create some "wide" data with columns corresponding to each year and rows corresponding to each municipality.

What we have right now are "long" data - where each observation represents a year and a municipality. To go wide to long and vice versa, there are some sophisticated functions in `dplyr` but we will do this in a simple way to start.

We need to split the data into 2001 and 2011 again and rename the columns.

We start by selecting observations from `summary` where `year == "2001"` using the `filter` command.

```{r filter_2000, echo=TRUE}
summarize2001  <- summary %>% 
  filter(year == "2001")
```

We do it again for 2011

```{r filter_2011, echo=TRUE}
summarize2011  <- summary %>% 
  filter(year == "2011")
```

Are `summarize2001` and `summarize2011` the same length? e.g. the same number of towns?

```{r summaries_same_length, echo=TRUE}
length(summarize2001) == length(summarize2011) 
```

We rename variables the `colnames` fucntion so that the field names in each summary data frame have years attached to them. 

```{r summary_colnames, echo=TRUE}
summarize2001 <- summarize2001 %>%
  rename(year_01 = year,
         MUNI_NAME_01 = MUNI_NAME,
         countPatch_01 = countPatch,
         mean_Compactness_01 = mean_Compactness,
         meanArea_01 = mean_Area,
         sum_Area_01 = sum_Area,
         sum_Perimter_01 = sum_Perimter)

summarize2011 <- summarize2011 %>%
  rename(year_11 = year,
         MUNI_NAME_11 = MUNI_NAME,
         countPatch_11 = countPatch,
         mean_Compactness_11 = mean_Compactness,
         meanArea_11 = mean_Area,
         sum_Area_11 = sum_Area,
         sum_Perimter_11 = sum_Perimter)
```

We use a series of tidy functions to bind the columns together from each summary data frame into one data frame using `cbind`. We use the `mutate` function from the `dplyr` package to create several new columns:

`MUNI_NAME` - A single MUNI Name unique ID which corresponds to our ArcGIS data set
`compactness_Change` - Compactness Change between 2001 and 2011, in percent 0-100
`sumAreaChange` - Sum of area change between 2001 and 2011, in percent 0-100
`numPatchChange` - Patch count change between 2001 and 2011, in percent 0-100

```{r create_summarize2, echo=TRUE}
summarize2 <- cbind(summarize2001, summarize2011) %>%
              as.data.frame() %>%
              mutate(MUNI_NAME = MUNI_NAME_01, # create one MUNI Name unique id
                    compactness_Change = ((mean_Compactness_11 - mean_Compactness_01) / mean_Compactness_01) * 100,
                    sumAreaChange = ((sum_Area_11 - sum_Area_01)/ sum_Area_01) *100, 
                    numPatchChange = ((countPatch_11 - countPatch_01) / countPatch_01) * 100) 
```

# 5. Create Maps

## 5.1 Import and join data

Import the Chester County boundary shapefile

```{r import_shp, echo=TRUE}
townsShapefile <- read_sf("~/Clients/MUSA_Teaching_and_Admin/CPLN_675_2020/wk_2/data/ChesterCounty_MuniBoundaries201107.shp")
```

Let's examine it - it's an `sf` object - like a data frame but with a geometry attached. Use the `View()` and `glimpse` commands to look at the tabular view of the data.

You can also do a quick look at it using the base plotting function in R

```{r plot_towns_shape, echo=TRUE}
plot(townsShapefile)
```

Next we add our summary data `summarize2` to the spatial data with a tabular join by `MUNI_NAME`.

This is a `left_join` where the `townsShapefile` is the left hand side and `summarize2` is on the right, meaning all the stuff on the left gets retained and joined only to elements of the right side that have a matching ID, in this case, `MUNI_NAME`.

Notice the result has as many observations as we have towns in `townsShapefile`.

```{r join_data_to_shp, echo=TRUE}
townsWithData <- left_join(townsShapefile, summarize2, by = "MUNI_NAME")
```

## 5.2. Create graphic styles

Before we map our data, we can create some graphic styles.

First a color palette of hexidecimal color codes which we can feed to our symbology options.

```{r color_ramp, echo=TRUE}
ramp8 <- c("#050234","#2C0078","#7F00BF","#F600FF","#FF0DBE","#FF569F","#FF9BA8","#FFE8E4")
```

Then we can set up a set of styles for our maps in ggplot/sf. We can call these any time we want to style a map. You can create your own themes depending on your preferences.

```{r mapTheme, echo=TRUE}
mapTheme <- theme(
    text = element_text( color = "black"),
    plot.title = element_text(size = 14,colour = "black"),
    plot.subtitle=element_text(face="italic"),
    plot.caption=element_text(hjust=0),
    axis.ticks = element_blank(),
    panel.background = element_blank(),axis.title = element_blank(),
    axis.text = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(colour = "black", fill=NA, size=2)
  ) 
```

## 5.3 Mapping change

First we can map the change in the total number of patches per town.

```{r patches_by_town, echo=TRUE}
ggplot() + 
  geom_sf(data=townsWithData, aes(fill=numPatchChange)) + 
  coord_sf() +
  scale_fill_gradientn(colors=ramp8, name = "Percent\nChange") +
  labs(title="Change in the number of undeveloped patches by town\nChester County (2001-2011)") +
  mapTheme
```  

We can also map the change in the area of those patches by town.

```{r patches_area_by_town, echo=TRUE} 
ggplot() + 
  geom_sf(data=townsWithData, aes(fill=sumAreaChange)) + 
  coord_sf() +
  scale_fill_gradientn(colors=ramp8, name = "Percent\nChange") +
  labs(title="Change in the area of non-developed patches\nChester County (2001-2011)") +
  mapTheme
```

Lastly we can map the change in compactness (isoperimetric quotient) during the span 2001-2011

```{r iso_by_town, echo=TRUE} 
ggplot() + 
  geom_sf(data=townsWithData, aes(fill=compactness_Change)) + 
  coord_sf() +
  scale_fill_gradientn(colors=ramp8, name = "Percent\nChange") +
  labs(title="Change in the undeveloped land fragmentation by town\nChester County (2001-2011)",
          caption="Fragmentation defined by the 'Isoperimetric quotient' - The compactness of a given landscape patch\nrelative to the most compact shape (a circle) with the same perimter. The lower the quotient, the \nmore fragmented a shape is considered. Fragmentation is related to a \ndecrease in the hardiness of ecosystems") +
  mapTheme
```


Raster function is aggregate